A general Nullstellensatz for generalized spaces

### 2019-05-10 slides-munich2019.tex
Slides for tomorrow's talk in Munich

\title{A general Nullstellensatz for generalised spaces}
\author{Ingo Blechschmidt}
\date{April 11th, 2019}

@@

\setbeamertemplate{frametitle}{%
  \vskip0.4em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
    \begin{tikzpicture}
      \def\R{8pt}
      \node (title) {\hil{\large\insertframetitle}};
      \begin{pgfonlayer}{background}
        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
      \end{pgfonlayer}
    \end{tikzpicture}
  \end{beamercolorbox}%
  \vskip-0.6em%
}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
\begin{frame}[c]
  \centering

  \bigskip
  \includegraphics[width=0.4\textwidth]{phantoms}
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
    \begin{pgfonlayer}{background}
      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

  \scriptsize
  \textit{-- an invitation --}
  \bigskip

  6th Workshop on Formal Topology in Birmingham \\
  April 11th, 2019
  \bigskip

  Ingo Blechschmidt \\
  Università di Verona
  \par
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 16.9cm 0,clip]{black-hole-m87}\end{minipage}}
\begin{frame}[plain]
  \vspace*{\paperheight}\vspace*{-0.8cm}\scriptsize
  \centering
  \textcolor{white}{Fig.: The Event Horizon Telescope picture of the central black
  hole in the galaxy M87}
  \par

  \jnote{1}{At this meeting we had lots of discussions on the notion of space.
  Here is a picture of actual space, or rather the lack thereof, released only
  yesterday.}
\end{frame}}
\addtocounter{framenumber}{-1}

@@

  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
  constructive but impredicative flavour of English (which could be formalised
  by what is supported by the internal language of elementary toposes with an NNO).
  However the Nullstellensatz presented later makes no use of the
  subobject classifier, hence the results can likely be generalised to hold in
  a predicative metatheory or to hold for arithmetic universes.}

@@

  \justifying
  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
  \hil{conservative} in that
  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
  \vspace*{-1.2em}
  \begin{enumerate}
    \item The sequent~$\sigma$ holds for~$U_\TT$.
    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
    \explainstub{topos}{purple!20}.
    \item The sequent~$\sigma$ is provable modulo~$\TT$.
  \end{enumerate}

@@

\begin{frame}{Construction of the generic model}
  The generic model is \hil{not} the same as \ldots
  \begin{itemize}
    \item the \hil{initial model} (think~$\ZZ$) or
    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
  \end{itemize}
  Set-based models are \hil{too inflexible}.

  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
  \begin{enumerate}
    \item objects: $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$ (shorter:
    $\{ \vec x\_ \varphi \}$)
    \item morphisms: eqv. classes of \explainstub{provably functional
    formulas}{red!20}
    \item coverings: \explainstub{provably jointly surjective
    families}{purple!30}
  \end{enumerate}
  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by~$\yon\{x\?X\_\top\}$.

@@

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$.}

  \jnote{2}{A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

  \vspace*{-0.6em}
  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
  \exists \vec x_i\_ \theta_i)$.

  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}

  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
  empty theory, that is, iff~$\TT$ has exactly one model in any topos.

  The special case that there exists at least some conservative~$\TT$-model
  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
  which is Morita-equivalent to the empty theory.}
\end{frame}

@@

\begin{frame}{Working internally to toposes}
  \justifying
  Let~$\C$ be a site. We recursively define
  \[
    U \models \varphi \quad
    \text{(``$\varphi$ holds on~$U$'')}
  \]
  for objects~$U \in \C$ and \squiggly{formulas}~$\varphi$.
  Write~``$\Sh(\C) \models \varphi$'' for~$1 \models \varphi$.
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ \ }c@{\ \ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ \text{the
  empty family is a covering of~$U$} \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $(U_i \to U)_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$V \to U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all $V \to U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  % U \models \forall F\_ \varphi(F) &\Ll&
  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$}
  %U \models \exists F\_ \varphi(F) &\Ll&
  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  %&&
  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]

@@

  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
  proves$^\star$~$\sigma$}{yellow!70}.}
  \end{quote}
  \vspace*{-0.6em}
}

@@

  \textbf{Remark.} Externalising the internal classifying
  topos~$\Set[\TT][\ul{\TT}/U_\TT]$ yields the classifying topos
  of~$\TT$-homomorphisms.

  \jnote{1}{
    Just as locales internal to a topos~$\E$ can be externalised to yield
    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
    externalised to yield bounded geometric morphisms. Since the composition of
    bounded geometric morphisms is bounded, the externalisation of a
    Grothendieck topos internally to a Grothendieck topos is itself a
    Grothendieck topos, hence the classifying topos of some geometric theory.

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves$^{\star\star}$~$\sigma$}. \end{equation}

  \textbf{Theorem 2.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:

  \vspace*{-0.4em}
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
  \end{quote}

  \textbf{Theorem 3.}
  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
  % context~$x\?X$, where any two such formulas are identified if and only
  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
  The morphism~$\mathrm{ev}$ below is an isomorphism.
  \begin{equation}\label{eq:hon}\tag{$\star$}
    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
  \end{equation}

  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme~($\star$).

@@

    Theorems~3 and~4 generalise the Nullstellensatz to the
    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.

@@

  \jnote{2}{
    Plans for the future include:
    \begin{itemize}\justifying
      \item Developing an Agda library for dealing with the internal language
      of toposes and related kinds of categories, employing Agda's
      meta-programming facilities; with such a library at hand,
      formalising the Nullstellensatz in Agda.
      \item Exploring Nullstellensatz-style results for arithmetic universes.
      \item Applying the Nullstellensatz in constructive algebra and algebraic
      geometry, along the lines of generic freeness and synthetic algebraic
      geometry. Most of the toposes in geometric use are actually uncharted
      territory from a logical point of view.
    \end{itemize}
  }
\end{frame}}

>>>>

\title{A general Nullstellensatz for generalized spaces}
\author{Ingo Blechschmidt}
\date{May 10th, 2019}

@@

\setbeamertemplate{frametitle}{%
  \vskip0.4em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
    \begin{tikzpicture}
      \def\R{8pt}
      \node (title) {\hil{\large\insertframetitle}};
    \end{tikzpicture}
  \end{beamercolorbox}%
  \vskip-0.6em%
}

@@

\begin{frame}[c]
  \centering

  \bigskip
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalized spaces}};
  \end{tikzpicture}

  \scriptsize
  \bigskip

  Logic Workshop in Munich \\
  May 10th, 2019
  \bigskip

  Ingo Blechschmidt \\
  Università di Verona
  \par
\end{frame}

@@

  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
  constructive but impredicative flavour of English (which could be formalized
  by what is supported by the internal language of elementary toposes with an NNO).
  However the Nullstellensatz presented later makes no use of the
  subobject classifier, hence the results can likely be generalized to hold in
  a predicative metatheory or to hold for arithmetic universes.}

@@

  \justifying
  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
  \hil{conservative} in that
  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
  \vspace*{-1.2em}
  \begin{enumerate}
    \item The sequent~$\sigma$ holds for~$U_\TT$.
    \item The sequent~$\sigma$ holds for any~(sheaf) model of~$\TT$.
    \item The sequent~$\sigma$ is provable modulo~$\TT$.
  \end{enumerate}

@@

\begin{frame}{On the generic model}
  The generic model is \hil{not} the same as \ldots
  \begin{itemize}
    \item the \hil{initial model} (think~$\ZZ$) or
    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
  \end{itemize}
  Set-based models are \hil{too inflexible}.

  The generic model is a \hil{sheaf model}.

@@

  In general, the generic model cannot be realized as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$.}

  \jnote{2}{The special case that the generic model of a theory~$\TT$ can be realized as
  a set-based model occurs iff~$\TT$ is Morita-equivalent to the
  empty theory, that is, iff~$\TT$ has exactly one model in any topos.

  The special case that there exists at least some conservative set-based~$\TT$-model
  occurs iff~$\TT$ has a conservative geometric expansion to a theory
  which is Morita-equivalent to the empty theory.}
\end{frame}

@@

\begin{frame}{A primer on sheaf semantics}
  \justifying
  \vspace*{-0.6em}
  Write~``$\Set[\TT] \models \varphi$'' for~$\{\_\top\} \models \varphi$. We
  define
  \[
    \{\vec x\_ \alpha\} \models \varphi \quad
    \text{(``$\varphi$ holds on stage~$\{\vec x\_\alpha\}$'')}
  \]
  for Horn formulas~$\alpha$ (in contexts~$\vec x$) and first-order
  formulas~$\varphi$.\par
  \vspace*{-1.2em}
  \footnotesize
  \[
  \renewcommand{\arraystretch}{1.08}\begin{array}{@{\!\!\!\!\!\!\!\!\!\!\!}l@{\ \ }c@{\ \ }l@{}}
  \{\vec x\_ \alpha\} \models \top &\Ll& \text{true} \\
  \{\vec x\_ \alpha\} \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\
  \alpha \seq{\vec x} \bot \\
  \{\vec x\_ \alpha\} \models s = t \? F &\Ll& s = t \in F(\{\vec x\_\alpha\}) \\
  \{\vec x\_ \alpha\} \models \varphi \wedge \psi &\Ll&
  \text{$\{\vec x\_ \alpha\} \models \varphi$ and $\{\vec x\_ \alpha\} \models \psi$} \\
  \{\vec x\_ \alpha\} \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$\{\vec x\_ \alpha\} \models \varphi$ or $\{\vec x\_ \alpha\} \models
  \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
  && \text{there exists a covering~$(\{ \vec y_i\_ \beta_i \} \xrightarrow{p_i} \{ \vec x\_ \alpha \})_i$} \\
  && \quad\quad\text{such that for all~$i$: $\{\vec y_i\_ \beta_i\} \models
  p_i^*\varphi$ or $\{\vec y_i\_ \beta_i\} \models p_i^*\psi$} \\
  \{\vec x\_ \alpha\} \models \varphi \Rightarrow \psi &\Ll&
  \text{for all~$\{\vec y\_\beta\} \xrightarrow{p} \{\vec x\_\alpha\}$: }
  \text{$\{\vec y\_\beta\} \models p^*\varphi$ implies $\{\vec y\_\beta\} \models
  p^*\psi$} \\
  \{\vec x\_ \alpha\} \models \forall s \? F\_ \varphi &\Ll&
  \text{for all $\{\vec y\_\beta\} \xrightarrow{p} \{\vec x\_\alpha\}$ and~$s_0
  \in F(\{\vec y\_\beta\})$: $\{\vec y\_\beta\} \models (p^*\varphi)[s_0/s]$} \\
  % U \models \forall F\_ \varphi(F) &\Ll&
  % \text{for all $V \to U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  \{\vec x\_ \alpha\} \models \exists s \? F\_ \varphi &\Ll&
  \hcancel{\text{there exists $s_0 \in F(\{\vec x\_\alpha\})$ such that $\{\vec
  x\_\alpha\} \models \varphi[s_0/s]$}}{0pt}{3pt}{0pt}{-2pt} \\
  &&
  \text{there exists a covering $(\{ \vec y_i\_ \beta_i \} \xrightarrow{p_i} \{
  \vec x\_ \alpha \})_i$ such that for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(\{\vec y_i\_\beta_i\})$ such that
  $\{\vec y_i\_\beta_i\} \models (p_i^*\varphi)[s_0/s]$}
  %U \models \exists F\_ \varphi(F) &\Ll&
  %\hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  %\varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  %&&
  %\text{there exists a covering $(U_i \to U)_i$ such that for all~$i$:} \\
  %&& \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]

@@

  \begin{quote}\emph{\mbox{For any \explainstub{geometric$^\star$ sequent~$\sigma$}{red!20} over
  the \explainstub{signature of~$\ul{\TT}/U_\TT$}{blue!20},}
  if~\explainstub{$\sigma$ holds for~$U_\TT$}{yellow!70}, then~\explainstub{$\ul{\TT}/U_\TT$
  proves~$\sigma$}{yellow!70}.}
  \end{quote}
  \vspace*{-0.6em}
}

@@

  \jnote{1}{
    Just as locales internal to a topos~$\E$ can be externalized to yield
    localic geometric morphisms into~$\E$, internal Grothendieck toposes can be
    externalized to yield bounded geometric morphisms. Since the composition of
    bounded geometric morphisms is bounded, the externalisation of a
    Grothendieck topos internally to a Grothendieck topos is itself a
    Grothendieck topos, hence the classifying topos of some geometric theory.

@@

\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves~$\sigma$}. \end{equation}

  \textbf{Theorem 2.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is a sheaf for the topology cutting out~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:

  \vspace*{-0.4em}
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}
  \end{quote}

  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme.

@@

    Theorems~3 and~4 generalize the Nullstellensatz to the
    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.

@@

  \jnote{2}{
    Plans for the future include:
    \begin{itemize}\justifying
      \item Developing an Agda library for dealing with the internal language
      of toposes and related kinds of categories, employing Agda's
      meta-programming facilities; with such a library at hand,
      formalising the Nullstellensatz in Agda.
      \item Exploring Nullstellensatz-style results for arithmetic universes.
      \item Applying the Nullstellensatz in constructive algebra and algebraic
      geometry, along the lines of generic freeness and synthetic algebraic
      geometry. Most of the toposes in geometric use are actually uncharted
      territory from a logical point of view.
    \end{itemize}
  }
\end{frame}
